---
title: "Costing Portal Data Preliminary Analysis"
author: "CIL Lab"
date: "`r Sys.Date()`" 
output: word_document
#output: pdf_document
#output: html_document
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Objective:
As a lab group, we developed some code to clean the costing portal dataset

# Background:
The initial dataset was a mess!

```{r, include= FALSE}
## setup the file structure so that this markdown document is in the main project folder along with the other folders noted below

DataSource <- "./data" # input raw dataset goes here
output.dir <-"./output" # output dataset writes to here
functions.dir <- "./functions" # directory for functions 

# any functions that are called up by this markdown document need to be sourced here
source(file.path(functions.dir,'FormatData.R'))
source(file.path(functions.dir,'SummarizeData.R'))
source(file.path(functions.dir,'SummarizeData_RBL.R'))
source(file.path(functions.dir,'PlotConservationActionCategories.R'))
## general functions that need to be read in. 
library(readr)
library(tidyverse)     ## install.packages("tidyverse")
library(data.table)
library(stringr)
library(knitr)

```

# Methods
First we standardized data fields for analysis. We did things like make assumptions about XXX, and define Y and Z.

```{r include=FALSE}
##this chunk is used to format the data set. 
## can use it to make settings etc.

FormatDataY <- "N" # set to Y if you want to run the format data function again

if (FormatDataY == "Y"){
RawData <- read_csv(paste(DataSource,"/CostPortal_20230918.csv", sep = "")) 
EditedData <- FormatData(RawData)  # make a function to format the data
} else {EditedData <- read_csv(paste(DataSource,"/EditedData.csv", sep = ""),show_col_types = FALSE)}

```

We then performed some summary statistics to better understand the dataset. 

Our dataset has `r dim(EditedData)[1]` rows and `r dim(EditedData)[2]` columns. There are `r dim(unique(EditedData[,3])[1])` years with studies, the oldest is from `r min(EditedData[,3],na.rm=T)`, the newest is from `r max(EditedData[,3],na.rm=T)`.

Specifically, we explored the number and proportion of studies that included cost data that involved protected area (PA) establishment or management costs, and also mention of the different categories of costs across the studies. 



```{r, include=FALSE}
# read in summary data
SummaryData <- SummarizeData(EditedData)
SummaryData_RBL <- SummarizeData_RBL(EditedData)
```


```{r results='asis'}
# make into a pretty table
kable(SummaryData_RBL)
kable(SummaryData)
```
We explored the relationship between how many years of cost data were collected for a study (its duration) and the time it was published.


```{r}
plot(EditedData$`Date Published`,EditedData$Duration,
     xlab = "Date Published", ylab = "Duration",
     main = "Duration of reported cost datasets over time")
```

We also explored which conservation actions were described by the studies and how they were distributed.

```{r results='asis'}
# try to return the plot
plot_cac <- PlotConservationActionCategories(EditedData)

```

We also explored how some of the numeric variables related to each other. This is useful for exploring predictor variables. For instance, if we wanted to formulate a regression exploring how the variation in how many years of cost data were reported in a study (duration) was related to the date it was published, whether capital and labor costs were included and whether it described protected area establishment costs (these are the numeric variables we have at the moment, no hypothesis as to why these particular predictors). 

```{r warning = FALSE, message = FALSE}

#### Correlation Matrix of Predictors vs Responses #######
## modified some code from http://handlesman.blogspot.com/2011/03/matrix-plot-with-confidence-intervals.html
### don't know what all of the settings are for, but it seems to be plotting alright.
# put histograms on the diagonal panel	
panel.hist <- function (x,...)					# define a function that says what we want to plot in the diagonal
{
  usr <- par("usr"); on.exit(par(usr))			# usr is setting the coordinates of a plot
  par(usr = c(usr[1:2],0,1.5))
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)			# make the hist 
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col="grey", ...)  # defines what the histogram is going to look like
}


## put correlations on the upper panels,
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y,use="everything")				
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  prefix <- "r = "
  rc <- cor.test(x,y,method = c("pearson"))				## calculate pearsons rho for upper grid
  txt <- paste(prefix,txt,sep="")
  text(0.5, 0.5, txt, cex = 1)
}

## plot a correlation matrix plot that uses the functions specified above to say what to plot where
pairs(EditedData[,c(30,3,32,34)],lower.panel = panel.smooth, diag.panel=panel.hist,upper.panel=panel.cor)



```

